Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2017-05-01 pkg/glmertree/paper/treatment_interactions.Rnw
discussion probably needs some more work...
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1299 255f4249-2263-4417-b1ea-8498779e27a7

We presented the GLMM tree algorithm, which allows for estimation of a GLM-based recursive partition, as well as estimation of random-effects parameters. We hypothesized GLMM tree to be well suited for the detection of treatment-subgroup interactions in clustered datasets, which was confirmed by our simulation studies.

GLMM tree accurately recovered the subgroups in 90\% of simulated datasets with treatment-subgroup interactions. In contrast, GLM tree accurately recovered treatment-subgroup interactions in only 61\% of these datasets. In terms of predictive accuracy, GLMM tree outperformed GLM tree as well as MERT, with an average correlation between true and predicted treatment-effect differences of .94 for GLMM tree, .88 for GLM tree and .75 for MERT. The Type-I error rate of GLMM tree very closely approximated the $\alpha$ level used for testing parameter stability: GLMM tree erroneously detected subgroups in 4\% of datasets without treatment-subgroup interactions, whereas GLM tree erroneously detected subgroups in 33\% of those datasets.

The better performance of GLMM tree was mostly observed when random effects in the datasets were sizable, and random intercepts were correlated with potential partitioning variables. In those datasets, GLM tree is likely to detect spurious splits and subgroups. At the same time, GLM tree showed less power to detect the true subgroups in the presence of random effects. As expecte, the accuracy of MERT was not affected much by the presence of random effects, but only approached the accuracy of GLMM tree in datasets with the largest sample and effect sizes. GLMM tree especially outperformed the other methods when effect size was small (i.e., Cohen's $d=.5$) and sample size was small (i.e., 200). Such effect and sample sizes are quite common in multi-center clinical trials, and GLMM tree may therefore provide a helpful tool for subgroup detection in those instances.

When random effects were absent from the simulated datasets, GLM and GLMM tree yielded very similar predictive accuracy. This finding is of practical importance, as it indicates that application of GLMM tree will not `hurt': GLM tree and GLMM tree are expected to perform equally well in the absence of random effects, while GLMM tree will likely outperform GLM tree in the presence of random effects.

Compared to linear mixed-effects models with pre-specified interactions, GLMM tree provided somewhat better accuracy, on average. As expected, GLMM tree performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at least partly piecewise. We found a clear advantage of GLMM tree in the presence of larger numbers of potential moderator variables, indicating that GLMM trees are much better suited than GLMMs for exploratory analyses. Also, GLMM trees may be easier to interpret: The number of terms in a GLMM increases quadratically with the number of potential moderator variables, yielding complex models. The trees in our simulations were limited to a maximum depth of three, requiring evaluation of at most 3 variables to make a prediction or decision in practice.

In the Application, we obtained similar findings: GLMM tree yielded higher predictive accuracy, while using a smaller number of variables for prediction than GLM tree and a GLMM with pre-specified interactions. In addition, the GLMM trees obtained over repeated subsamples of the training data proved to be relatively stable.

@@

Recursive partitioning methods were originally developed as a non-parametric tool for classification and regression, assuming the mechanism that generated the data unknown \cite<e.g., >{Brei01TwoCult}. However, GLMM tree is obviously a parametric tool, as it fits fixed-effects linear models in the nodes of the tree and a global model for the random effects, in turn introducing several distributional assumptions about the random effects and errors. Misspecification of these distributions will likely have a negative effect on the accuracy of the estimated GLMM tree.

Furthermore, misspecification of partitioning and fixed- and random-effects variables will also reduce accuracy of the resulting GLMM tree. If relevant variables are omitted, or incorrectly specified, GLMM tree can only approximate the true subgroups using the specified variables. Our simulations indicate that LM tree detected spurious subgroups as a result of misspecifying (that is, not including) the random effects. Reduced accuracy and spurious splits can also be expected to occur when relevant random effects are not included when specifying the GLMM tree model. Furthermore, as the random effects are estimated globally, misspecification of the random effects can have a strong impact on the resulting GLMM tree model.

>>>>

\readme{This summary is quite extensive. Personally, I would prefer a more condensed summary of `highlights'.}

We presented the GLMM tree algorithm, which allows for estimation of a GLM-based recursive partition, as well as estimation of random-effects parameters. We hypothesized GLMM trees to be well suited for the detection of treatment-subgroup interactions in clustered datasets and confirmed this by our simulation studies.

GLMM trees accurately recovered the subgroups in 90\% of simulated datasets with treatment-subgroup interactions. In contrast, GLM trees accurately recovered treatment-subgroup interactions in only 61\% of these datasets. In terms of predictive accuracy, GLMM trees outperformed GLM trees as well as MERT, with an average correlation between true and predicted treatment-effect differences of .94 for GLMM trees, .88 for GLM trees and .75 for MERT. The Type-I error rate of GLMM trees very closely approximated the $\alpha$ level used for testing parameter stability: GLMM trees erroneously detect subgroups in 4\% of datasets without treatment-subgroup interactions, whereas GLM trees erroneously detect subgroups in 33\% of those datasets.

The better performance of GLMM trees was mostly observed when random effects in the datasets were sizeable, and random intercepts were correlated with potential partitioning variables. In those datasets, GLM trees are likely to detect spurious splits and subgroups. At the same time, GLM trees showed less power to detect the true subgroups in the presence of random effects. As expected, the accuracy of MERT was not affected much by the presence of random effects, but only approached the accuracy of GLMM trees in datasets with the largest sample and effect sizes. GLMM trees especially outperformed the other methods when effect size was small (i.e., Cohen's $d=.5$) and sample size was small (i.e., 200). Such effect and sample sizes are quite common in multi-center clinical trials, and GLMM trees may therefore provide a helpful tool for subgroup detection in those instances.

When random effects were absent from the simulated datasets, GLM and GLMM trees yielded very similar predictive accuracy. This finding is of practical importance, as it indicates that application of GLMM trees will not `hurt': GLM trees and GLMM trees are expected to perform equally well in the absence of random effects, while GLMM trees will likely outperform GLM trees in the presence of random effects.

Compared to linear mixed-effects models with pre-specified interactions, GLMM trees provided somewhat better accuracy, on average. As expected, GLMM trees performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at least partly piecewise. We found a clear advantage of GLMM trees in the presence of larger numbers of potential moderator variables, indicating that GLMM trees are much better suited than GLMMs for exploratory analyses. Also, GLMM trees may be easier to interpret: The number of terms in a GLMM increases quadratically with the number of potential moderator variables, yielding complex models. The trees in our simulations were limited to a maximum depth of three, requiring evaluation of at most 3 variables to make a prediction or decision in practice.

In the Application, we obtained similar findings: the GLMM tree yielded higher predictive accuracy, while using a smaller number of variables for prediction than the GLM tree and a GLMM with pre-specified interactions. In addition, the GLMM trees obtained over repeated subsamples of the training data proved to be relatively stable.

@@

\readme{I would emphasize early on that GLMM trees rely on the properties of their building blocks, namely GLMMs and MOBs. The effect of misspecifications are studied to some degree for both methods. For GLMM one could cite papers discussing this. For MOB, or trees more broadly, there are also some references. One of these could be the `To Split or to Mix?' paper by Frick, Strobl, and Zeileis.}

Recursive partitioning methods were originally developed as a non-parametric tool for classification and regression, assuming the mechanism that generated the data unknown \cite<e.g., >{Brei01TwoCult}. However, GLMM trees are obviously a parametric tool, as they fit fixed-effects linear models in the nodes of the tree and a global model for the random effects, in turn introducing several distributional assumptions about the random effects and errors. Misspecification of these distributions will likely have a negative effect on the accuracy of the estimated GLMM tree.

Furthermore, misspecification of partitioning and fixed- and random-effects variables will also reduce accuracy of the resulting GLMM tree. If relevant variables are omitted, or incorrectly specified, GLMM tree can only approximate the true subgroups using the specified variables. Our simulations indicate that LM trees detect spurious subgroups as a result of misspecifying (that is, not including) the random effects. Reduced accuracy and spurious splits can also be expected to occur when relevant random effects are not included when specifying the GLMM tree model. Furthermore, as the random effects are estimated globally, misspecification of the random effects can have a strong impact on the resulting GLMM tree model.
